Towards a Verified Tableau Prover for Multi-Level-Syllogistics

### 2022-09-14 paper/paper.tex
Continue

\begin{abstract}
\end{abstract}

@@

The unquantified fragment \MLSS{} of set theory is then formed by combining set atoms, which are injected into formulae with the constructor \lstinline!&\fmAtom&!, with propositional logical connectives.
\todo[inline]{Cite Nipkow's Linear Quantifier Elimination?}
\begin{lstlisting}
datatype (atoms: 'a) fm =
  &\fmAtom& 'a | &\fmNeg{('a fm)}&
| &\fmAnd{'a fm}{'a fm}& | &\fmOr{'a fm}{'a fm}&

type_synonym 'a pset_fm = 'a pset_atom fm
\end{lstlisting}
We will often drop the atom constructor \lstinline!&\fmAtom{}&! to reduce clutter.
Additionally, we use the abbreviations \lstinline!s $\notinS$ t! and \lstinline!s $\neqS$ t! to denote \lstinline!&\fmNeg{\fmAtom{} (s $\inS$ t)}&! and \lstinline!&\fmNeg{\fmAtom{} (s $\eqS$ t)}&!, respectively.
Note that Isabelle's datatype package automatically defines a function \lstinline!atoms :: 'a fm $\Rightarrow$ 'a set! that allows us to obtain all set atoms in a formula.
We combine these functions in order to define \lstinline!vars! that extracts all the variable names occuring in a set formula.
\todo[inline]{Motivation: one rule introduces new variables, Mention vars polymorphic}
\begin{lstlisting}
definition vars :: 'a pset_fm $\Rightarrow$ 'a set where
  vars $\phi$ $\equiv$ $\bigcup$(vars ` atoms $\phi$)
\end{lstlisting}
In addition to the variable names that appear in a formula, we will later also need to obtain those set terms that occur as subterms in a formula. 
We first define a function that collects the subterms of a set term.
\todo[inline]{Motivation: linear rules don't introduce subterms, \\ \lstinline!subterms b $\approx$ subterms (last b)!}
\begin{lstlisting}
fun subterms :: 'a pset_term
             $\Rightarrow$ 'a pset_term set where
  subterms $\emptyset$ = {$\emptyset$}
| subterms (Var x) = {Var x}
| subterms (s $\unionS$ t) =
    {s $\unionS$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (s $\interS$ t) =
    {s $\interS$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (s $\diffS$ t) =
    {s $\diffS$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (Single t) =
    {Single t} $\cup$ subterms t
\end{lstlisting}
The subterms of an atom are then the subterms of those terms that make up the atom.
\begin{lstlisting}
fun subterms :: 'a pset_atom
             $\Rightarrow$ 'a pset_term set where
  subterms (s $\inS$ t) = subterms s $\cup$ subterms t
| subterms (s $\eqS$ t) = subterms s $\cup$ subterms t
\end{lstlisting}
Ultimately, we lift this to formulae by using \lstinline!atoms! again.
\begin{lstlisting}
definition subterms ::
    'a pset_fm $\Rightarrow$ 'a pset_term set where
  subterms $\phi$ $\equiv$ $\bigcup$(subterms ` atoms $\phi$)
\end{lstlisting}
The last function that we define computes the subformulae of a formula.
\begin{lstlisting}
fun subfms :: 'a fm $\Rightarrow$ 'a fm set where
  subfms (&\fmAtom& a) = {&\fmAtom& a}
| subfms (&\fmAnd{p}{q}&) =
    {&\fmAnd{p}{q}&} $\cup$ subfms p $\cup$ subfms q
| subfms (&\fmOr{p}{q}&) =
    {&\fmOr{p}{q}&} $\cup$ subfms p $\cup$ subfms q
| subfms (&\fmNeg{q}&) = {&\fmNeg{q}&} $\cup$ subfms q
\end{lstlisting}

@@

definition mlss_proc :: 'a pset_fm $\Rightarrow$ bool where
  mlss_proc $\phi$ $\equiv$ mlss_proc_branch [$\phi$]
\end{lstlisting}
\caption{Definition of the decision procedure through the \lstinline|mlss_proc_branch| and \lstinline|mlss_proc|.\label{fig:alg}}
\end{figure*}

@@

\subsection{Characterisation of the Pure Witnesses\label{sec:pwits}}
Recall that the pure witnesses for a branch \lstinline!b! are those witnesses that are not related to other subterms in \lstinline!last b! by equality.
In the context of a well-formed branch, this characterisation can be strengthened to any set term and, in addition, we also get that there is no membership literal where a pure witness is on the right-hand side.
Intuitively speaking, the realization of a pure witness does not depend on the realization of any other set term.
\begin{lstlisting}
lemma lemma_2:
  assumes wf_branch b
  assumes c $\in$ wits' b
  shows (Var c $\eqS$ t) $\notin$ set b
    and (t $\eqS$ Var c) $\notin$ set b
    and (t $\inS$ Var c) $\notin$ set b
\end{lstlisting}
So why are pure witnesses treated differently?
According to the definition of \lstinline!realize!, the pure witnesses would be evaluated to the empty set \lstinline!0 :: V!, were they not treated separately. 
To see that this is a problem, consider the branch
\begin{lstlisting}
  b = [Var s $\neqS$ Var t, Var t $\neqS$ Var u]
\end{lstlisting}
which expands to several open and saturated branches, one of which is
\begin{lstlisting}
  [Var x $\inS$ Var s, Var x $\notinS$ Var t,
   Var y $\inS$ Var t, Var y $\notinS$ Var u] @ b
\end{lstlisting}
for some fresh \lstinline!x! and \lstinline!y!.
Assigning both \lstinline!Var x! and \lstinline!Var y! a value of \lstinline!0! would result in the realization of \lstinline!Var s! and \lstinline!Var t! becoming equal;
therefore, the realization would not be a model for \lstinline!Var s $\neqS$ Var t!.
To prevent this, we assign the pure witnesses pairwise different values.

\begin{lstlisting}
fun is_literal :: 'a fm $\Rightarrow$ bool where
  is_literal (&\fmAtom& _) = True
| is_literal (&\fmNeg{(\fmAtom{} \_)}&) = True
| is_literal _ = False
\end{lstlisting}

\begin{lstlisting}
definition no_new_subterms b $\equiv$
  $\forall$t $\in$ subterms b. t $\notin$ Var ` wits b
    $\longrightarrow$ t $\in$ subterms (last b))
\end{lstlisting}

\subsection{Realisation of an Open Branch\label{sec:real}}
Remember that for completeness, we need to show that the realization function for an open and saturated branch \lstinline!b! actually constitutes a model for all formulae in the branch.
We start by verifying that the realization function models all literals in the branch; more formally, the following propositions hold: 
\begin{enumerate}[label=(\roman*)]
  \item\label{it:at_mem} We have \lstinline!realize s $\in$ elts (realize t)! if it holds that \lstinline!s $\inS$ t! is in \lstinline!b!. 
  \item\label{it:at_eq} We have \lstinline!realize s $=$ realize t! if \lstinline!s $\eqS$ t! is in \lstinline!b!.
  \item\label{it:af_eq} We have \lstinline!realize s $\neq$ realize t! if \lstinline!s $\neqS$ t! is in \lstinline!b!.
  \item\label{it:af_mem} We have \lstinline!realize s $\notin$ elts (realize t)! if it holds that \lstinline!s $\notinS$ t! is in \lstinline!b!. 
\end{enumerate}
To illustrate the usefulness of \lstinline!lemma_2!, we prove Proposition~\ref{it:at_eq}.
The proofs of all propositions translate well into Isabelle so we refer to the original paper~\cite{new_fast_tableau} for the remaining proofs. 
\begin{proof}[Proof of Proposition~\ref{it:at_eq}]
  Assume that \lstinline!s $\eqS$ t! is in \lstinline!b!.
  If there exists a \lstinline!c $\in$ wits' b! where \lstinline!s = Var c! or \lstinline!t = Var c!, we arrive at a contradiction due to \lstinline!lemma_2!.
  Therefore, both \lstinline!s $\in$ subterms' b! and \lstinline!t $\inS$ subterms' b! must hold.
  Now, assume for contradiction that
\begin{lstlisting}
  &\phantom{$\Longrightarrow$}& realize s $\neq$ realize t
  $\Longrightarrow$ elts (realize s) $\neq$ elts (realize t).
\end{lstlisting}
  Without loss of generality --- the other case is symmetric ---, we obtain an \lstinline!a! such that
\begin{lstlisting}
  a $\in$ elts (realize s) $\land$ a $\notin$ elts (realize t). 
\end{lstlisting}
  Considering the fact that \lstinline!s $\in$ subterms' b! and the definition of \lstinline!realize!, we obtain an \lstinline!u! such that \lstinline!a = realize u! and \lstinline!u $\rightarrow_\text{bgraph b}$ s!.
  This, in turn, yields that \lstinline!u $\inS$ s! must be in \lstinline!b!.
  Together with the assumption \lstinline!(s $\eqS$ t) $\in$ b! and the openness of \lstinline!b!, it follows that \lstinline!u $\inS$ t! must also be in \lstinline!b!.
  But then we have
\begin{lstlisting}[columns=fixed]
  realize u $\in$ elts (realize t)
        = a $\in$ elts (realize t)
\end{lstlisting}
  using Proposition~\ref{it:at_mem}, which stands in contradiction to the assumption \lstinline!a $\notin$ elts (realize t)!.
\end{proof}

\noindent The results on literals can now be lowered to set terms.
\begin{enumerate}[label=(\alph*)]
  \item\label{it:empty} If \lstinline!$\emptyset$ $\in$ subterms b!, then \lstinline!realize $\emptyset$ = 0!.
  \item\label{it:op} Let \lstinline!$\star_\text{s}$ $\in$ {$\unionS$, $\diffS$, $\interS$}!. If the term \lstinline!s $\star_\text{s}$ t! occurs in \lstinline!subterms b!, then
    \begin{center}
    \lstinline!realize (s $\star_\text{s}$ t) = realize s $\star$ realize t!.
    \end{center}
  \item\label{it:single} If \lstinline!Single t $\in$ subterms b!, then
    \begin{center}
    \lstinline!realize (Single t) = vset {realize t}!.
    \end{center}
\end{enumerate}
\begin{proof}[Proof of Proposition~\ref{it:empty}]
  Since \lstinline!b! is open, we know \lstinline!s $\inS$ $\emptyset$! is not in \lstinline!b! for any \lstinline!s!, implying that the term \lstinline!$\emptyset$! has no parents in \lstinline!bgraph b!.
  Moreover, the assumption \lstinline!$\emptyset$ $\in$ subterms b! can be strengthened to \lstinline!$\emptyset$ $\in$ subterms' b! as the term \lstinline!$\emptyset$! is not a witness. 
  Ultimately, we obtain \lstinline!realize $\emptyset$ = 0! by definition of \lstinline!realize!.
\end{proof}
\begin{proof}[Proof of Proposition~\ref{it:op} for $\star_\texttt{s} = -_\texttt{s}$]
  For equality, it suffices to prove subset inclusion in both directions.

  \textbf{Direction $\subseteq$. } Fix some \lstinline!e! in \lstinline!elts (realize (s $\diffS$ t))!.
  Again, we obtain an \lstinline!u! with 
\begin{lstlisting}
  e = realize u $\land$ u $\rightarrow_\text{bgraph b}$ (s $\diffS$ t).
\end{lstlisting}
  We therefore have \lstinline!(u $\inS$ s $\diffS$ t) $\in$ set b! which, due to \lstinline!b! being saturated, implies that \lstinline!u $\inS$ s! is in \lstinline!b! while \lstinline!u $\inS$ t! is not.
  By applying Proposition~\ref{it:at_mem} and~\ref{it:af_mem} we obtain
\begin{lstlisting}
  realize u $\in$ realize s $\land$ realize u $\notin$ realize t
\end{lstlisting}
  and together with \lstinline!e = realize u! we arrive at the goal 
\begin{lstlisting}
  e $\in$ elts (realize s - realize t).
\end{lstlisting}

@@

\noindent The final step towards obtaining a proper model is to connect the realization function to the semantics as defined in \autoref{sec:semantics}.
For set terms, we can use the Propositions~\ref{it:empty}--\ref{it:single} to prove the lemma below by induction on \lstinline!t!.
\begin{lstlisting}
lemma
  assumes t $\in$ subterms b
  shows &\Ist& ($\lambda$x. realize (Var x)) t = realize t
\end{lstlisting}
Lifting the above result to formulae yields the coherence of \lstinline!b! as the original paper~\cite{new_fast_tableau} calls it.
The proof is a tedious but straightforward induction on the the size of the formulae.
\begin{lstlisting}
lemma coherence:
  assumes $\phi$ $\in$ set b
  shows ($\lambda$x. realize (Var x)) $\models$ $\phi$
\end{lstlisting}
The coherence property finishes the proof of completeness of the calculus as it gives us a model for every formula in an open and saturated branch.

\section{Soundness of the Calculus\label{sec:sound}}
Two aspects determine the soundness of the calculus: a closed branch must be unsatisfiable and all expansion rules must maintain satisfiability.
We can express the former aspect in Isabelle as follows.
\begin{lstlisting}[belowskip=0pt]
lemma bclosed_sound:
  assumes bclosed b shows $\exists$$\phi$ $\in$ set b. M $\not\models$ $\phi$
\end{lstlisting}
\begin{proof}
  Assume for contradiction that \lstinline!b! is closed but we have a model, i.e.\ it holds that \lstinline!$\forall$$\phi$ $\in$ set b. M $\models$ $\phi$!.
  We perform a case analysis on \lstinline!bclosed b!.
  It is clear that, for any \lstinline!s!, neither does \lstinline!M! model \lstinline!s $\in$ $\emptyset$! nor \lstinline!s $\neqS$ s!.
  Furthermore, no model can satisfy both \lstinline!$\phi$! and \lstinline!&\fmNeg{$\phi$}&! at the same time.
  This leaves us with the case where \lstinline!b! contains a membership cycle
\begin{lstlisting}
  t$_\text{0}$ $\inS$ t$_\text{1}$, t$_\text{1}$ $\inS$ t$_\text{2}$, $\ldots$, t$_\text{k}$ $\inS$ t$_\text{0}$.
\end{lstlisting}
  By assumption, each atom of the membership cycle holds under the model \lstinline!M!; in other words, we have
\begin{lstlisting}
  &\Ist& M t$_\text{i}$ $\in$ elts (&\Ist& M t$_\text{i + 1 mod k + 1}$)
\end{lstlisting}
  for each \lstinline!i $\in$ {0,$\ldots$,k}!.
  Therefore, the transitive closure of the membership relation \lstinline!{(x, y). x $\in$ elts y}! contains a cycle. 
  This, however, is a contradiction since the membership relation of \lstinline!V! is well-founded (see \autoref{sec:semantics}) and well-foundedness is closed under transitive closure.
\end{proof}

>>>>

\begin{abstract}
  Using Isabelle/HOL, we verify the state-of-the-art decision procedure for multi-level syllogistics with singleton (\MLSS{} for short), which is a ground fragment of set theory.
  We formalise the syntax and semantics of \MLSS{} as well as a sound and complete tableau calculus for it.
  We also provide an abstract specification of a decision procedure that applies the rules of the calculus exhaustively and prove its termination.
\end{abstract}

@@

The unquantified fragment \MLSS{} of set theory is then formed by combining set atoms, which are injected into formulae with the constructor \lstinline!&\fmAtom&!, with propositional logical connectives.
\todo[inline]{Cite Nipkow's Linear Quantifier Elimination?, literals}
\begin{lstlisting}
datatype (atoms: 'a) fm =
  &\fmAtom& 'a | &\fmNeg{('a fm)}&
| &\fmAnd{'a fm}{'a fm}& | &\fmOr{'a fm}{'a fm}&

type_synonym 'a pset_fm = 'a pset_atom fm
\end{lstlisting}

\begin{lstlisting}
fun is_literal :: 'a fm $\Rightarrow$ bool where
  is_literal (&\fmAtom& _) = True
| is_literal (&\fmNeg{(\fmAtom{} \_)}&) = True
| is_literal _ = False
\end{lstlisting}
We will often drop the atom constructor \lstinline!&\fmAtom{}&! to reduce clutter.
Additionally, we use the abbreviations \lstinline!s $\notinS$ t! and \lstinline!s $\neqS$ t! to denote \lstinline!&\fmNeg{\fmAtom{} (s $\inS$ t)}&! and \lstinline!&\fmNeg{\fmAtom{} (s $\eqS$ t)}&!, respectively.
Note that Isabelle's datatype package automatically defines a function \lstinline!atoms :: 'a fm $\Rightarrow$ 'a set! that allows us to obtain all set atoms in a formula.
We combine these functions in order to define \lstinline!vars! that extracts all the variable names occuring in a set formula.
\todo[inline]{Motivation: one rule introduces new variables, Mention vars polymorphic}
\begin{lstlisting}
definition vars :: 'a pset_fm $\Rightarrow$ 'a set where
  vars $\phi$ $\equiv$ $\bigcup$(vars ` atoms $\phi$)
\end{lstlisting}
In addition to the variable names that appear in a formula, we will later also need to obtain those set terms that occur as subterms in a formula. 
We first define a function that collects the subterms of a set term.
\todo[inline]{Motivation: linear rules don't introduce subterms, \\ \lstinline!subterms b $\approx$ subterms (last b)!}
\begin{lstlisting}
fun subterms :: 'a pset_term
             $\Rightarrow$ 'a pset_term set where
  subterms $\emptyset$ = {$\emptyset$}
| subterms (Var x) = {Var x}
| subterms (s $\unionS$ t) =
    {s $\unionS$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (s $\interS$ t) =
    {s $\interS$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (s $\diffS$ t) =
    {s $\diffS$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (Single t) =
    {Single t} $\cup$ subterms t
\end{lstlisting}
The subterms of an atom are then the subterms of those terms that make up the atom.
\begin{lstlisting}
fun subterms :: 'a pset_atom
             $\Rightarrow$ 'a pset_term set where
  subterms (s $\inS$ t) = subterms s $\cup$ subterms t
| subterms (s $\eqS$ t) = subterms s $\cup$ subterms t
\end{lstlisting}
Ultimately, we lift this to formulae by using \lstinline!atoms! again.
\begin{lstlisting}
definition subterms ::
    'a pset_fm $\Rightarrow$ 'a pset_term set where
  subterms $\phi$ $\equiv$ $\bigcup$(subterms ` atoms $\phi$)
\end{lstlisting}
The last function that we define computes the subformulae of a formula.
\begin{lstlisting}
fun subfms :: 'a fm $\Rightarrow$ 'a fm set where
  subfms (&\fmAtom& a) = {&\fmAtom& a}
| subfms (&\fmAnd{p}{q}&) =
    {&\fmAnd{p}{q}&} $\cup$ subfms p $\cup$ subfms q
| subfms (&\fmOr{p}{q}&) =
    {&\fmOr{p}{q}&} $\cup$ subfms p $\cup$ subfms q
| subfms (&\fmNeg{q}&) = {&\fmNeg{q}&} $\cup$ subfms q
\end{lstlisting}

@@

definition mlss_proc :: 'a pset_fm $\Rightarrow$ bool where
  mlss_proc $\phi$ $\equiv$ mlss_proc_branch [$\phi$]
\end{lstlisting}
\caption{Definition of the decision procedure through \lstinline|mlss_proc_branch| and \lstinline|mlss_proc|.\label{fig:alg}}
\end{figure*}

@@

\subsection{Characterisation of the Pure Witnesses\label{sec:pwits}}
Recall that the pure witnesses for a branch \lstinline!b! are those witnesses that are not related to other subterms in \lstinline!last b! by equality.
In the context of a well-formed branch, this characterisation can be strengthened to any set term and, in addition, we also get that there is no membership literal where a pure witness is on the right-hand side.
Intuitively speaking, the realization of a pure witness does not depend on the realization of any other set term.
\begin{lstlisting}[label={lst:lemma_2}]
lemma lemma_2:
  assumes wf_branch b
  assumes c $\in$ wits' b
  shows (Var c $\eqS$ t) $\notin$ set b
    and (t $\eqS$ Var c) $\notin$ set b
    and (t $\inS$ Var c) $\notin$ set b
\end{lstlisting}
So why are pure witnesses treated differently?
According to the definition of \lstinline!realize!, the pure witnesses would be evaluated to the empty set \lstinline!0 :: V!, were they not treated separately. 
To see that this is a problem, consider the branch
\begin{lstlisting}
  b = [Var s $\neqS$ Var t, Var t $\neqS$ Var u]
\end{lstlisting}
which expands to several open and saturated branches, one of which is
\begin{lstlisting}
  [Var x $\neqS$ Var y,
   Var x $\inS$ Var s, Var x $\notinS$ Var t,
   Var y $\inS$ Var t, Var y $\notinS$ Var u] @ b
\end{lstlisting}
for some fresh \lstinline!x! and \lstinline!y!.
Assigning both \lstinline!Var x! and \lstinline!Var y! a value of \lstinline!0! would contradict the literal \lstinline!Var x $\neqS$ Var y!.
To prevent this, we assign the pure witnesses pairwise different values.

The proof of \hyperref[lst:lemma_2]{\lstinline!lemma_2!} is more technical than interesting so we refer the reader to the formalisation.

\subsection{Realisation of an Open Branch\label{sec:real}}
Remember that for completeness, we need to show that the realization function for an open and saturated branch \lstinline!b! actually constitutes a model for all formulae in the branch.
We start by verifying that the realization function models all literals in the branch; more formally, the following propositions hold: 
\begin{enumerate}[label=(\roman*)]
  \item\label{it:at_mem} We have \lstinline!realize s $\in$ elts (realize t)! if it holds that \lstinline!s $\inS$ t! is in \lstinline!b!. 
  \item\label{it:at_eq} We have \lstinline!realize s $=$ realize t! if \lstinline!s $\eqS$ t! is in \lstinline!b!.
  \item\label{it:af_eq} We have \lstinline!realize s $\neq$ realize t! if \lstinline!s $\neqS$ t! is in \lstinline!b!.
  \item\label{it:af_mem} We have \lstinline!realize s $\notin$ elts (realize t)! if it holds that \lstinline!s $\notinS$ t! is in \lstinline!b!. 
\end{enumerate}
To illustrate the usefulness of \hyperref[lst:lemma_2]{\lstinline!lemma_2!}, we prove Proposition~\ref{it:at_eq}.
The proofs of all propositions translate well into Isabelle so we refer to the original paper~\cite{new_fast_tableau} for the remaining proofs. 
\begin{proof}[Proof of Proposition~\ref{it:at_eq}]
  Assume that \lstinline!s $\eqS$ t! is in \lstinline!b!.
  If there exists a \lstinline!c $\in$ wits' b! where \lstinline!s = Var c! or \lstinline!t = Var c!, we arrive at a contradiction due to \hyperref[lst:lemma_2]{\lstinline!lemma_2!}.
  Therefore, both \lstinline!s $\in$ subterms' b! and \lstinline!t $\inS$ subterms' b! must hold.
  Now, assume for contradiction that
\begin{lstlisting}
  &\phantom{$\Longrightarrow$}& realize s $\neq$ realize t
  $\Longrightarrow$ elts (realize s) $\neq$ elts (realize t).
\end{lstlisting}
  Without loss of generality --- the other case is symmetric ---, we obtain an \lstinline!a! such that
\begin{lstlisting}
  a $\in$ elts (realize s) $\land$ a $\notin$ elts (realize t). 
\end{lstlisting}
  Considering the fact that \lstinline!s $\in$ subterms' b! and the definition of \lstinline!realize!, we obtain an \lstinline!u! such that \lstinline!a = realize u! and \lstinline!u $\rightarrow_\text{bgraph b}$ s!.
  This, in turn, yields that \lstinline!u $\inS$ s! must be in \lstinline!b!.
  Together with the assumption \lstinline!(s $\eqS$ t) $\in$ b! and the openness of \lstinline!b!, it follows that \lstinline!u $\inS$ t! must also be in \lstinline!b!.
  But then we have
\begin{lstlisting}[columns=fixed]
  realize u $\in$ elts (realize t)
        = a $\in$ elts (realize t)
\end{lstlisting}
  using Proposition~\ref{it:at_mem}, which stands in contradiction to the assumption \lstinline!a $\notin$ elts (realize t)!.
\end{proof}

\noindent The results on literals can now be lowered to set terms.
\begin{enumerate}[label=(\alph*)]
  \item\label{it:empty} It holds that \lstinline!realize $\emptyset$ = 0!.
  \item\label{it:op} Let \lstinline!$\star_\text{s}$ $\in$ {$\unionS$, $\diffS$, $\interS$}!. If the term \lstinline!s $\star_\text{s}$ t! occurs in \lstinline!subterms b!, then
    \begin{center}
    \lstinline!realize (s $\star_\text{s}$ t) = realize s $\star$ realize t!.
    \end{center}
  \item\label{it:single} If \lstinline!Single t $\in$ subterms b!, then
    \begin{center}
    \lstinline!realize (Single t) = vset {realize t}!.
    \end{center}
\end{enumerate}
\begin{proof}[Proof of Proposition~\ref{it:empty}]
  By definition, \lstinline!realize $\emptyset$! is equal to \lstinline!0! if \lstinline!$\emptyset$! is not a vertex of \lstinline!bgraph b!.
  Otherwise, \lstinline!$\emptyset$! must be in \lstinline!subterms' b! as it is not a witness.
  Moreover, since \lstinline!b! is open, we know \lstinline!s $\inS$ $\emptyset$! is not in \lstinline!b! for any \lstinline!s!, implying that the term \lstinline!$\emptyset$! has no parents in \lstinline!bgraph b!.
  Ultimately, we obtain \lstinline!realize $\emptyset$ = 0!.
\end{proof}
\begin{proof}[Proof of Proposition~\ref{it:op} for $\star_\texttt{s} = -_\texttt{s}$]
  For equality, it suffices to prove subset inclusion in both directions.

  \textbf{Direction $\subseteq$. } Fix some \lstinline!e! in \lstinline!elts (realize (s $\diffS$ t))!.
  Again, we obtain an \lstinline!u! with 
\begin{lstlisting}
  e = realize u $\land$ u $\rightarrow_\text{bgraph b}$ (s $\diffS$ t).
\end{lstlisting}
  We therefore have \lstinline!(u $\inS$ s $\diffS$ t) $\in$ set b! which, due to \lstinline!b! being saturated, implies that \lstinline!u $\inS$ s! and \lstinline!u $\notinS$ t! are in \lstinline!b!.
  \todo[inline]{Explain latter of line above}
  By applying Proposition~\ref{it:at_mem} and~\ref{it:af_mem} we obtain
\begin{lstlisting}
  realize u $\in$ realize s $\land$ realize u $\notin$ realize t
\end{lstlisting}
  and together with \lstinline!e = realize u! we arrive at the goal 
\begin{lstlisting}
  e $\in$ elts (realize s - realize t).
\end{lstlisting}

@@

\noindent The final step towards obtaining a proper model is to connect the realization function to the semantics as defined in \autoref{sec:semantics}.
For set terms, we can use the Propositions~\ref{it:empty}--\ref{it:single} to prove the lemma below by induction on \lstinline!t!.
\begin{lstlisting}
lemma
  assumes t $\in$ subterms b
  shows &\Ist& ($\lambda$x. realize (Var x)) t = realize t
\end{lstlisting}
Lifting the above result to formulae yields the coherence of \lstinline!b! as the original paper~\cite{new_fast_tableau} calls it.
The proof is a tedious but straightforward induction on the the size of the formulae.
\begin{lstlisting}[label={lst:coherence}]
lemma coherence:
  assumes $\phi$ $\in$ set b
  shows ($\lambda$x. realize (Var x)) $\models$ $\phi$
\end{lstlisting}
The coherence property finishes the proof of completeness of the calculus as it gives us a model for every formula in an open and saturated branch.

\section{Soundness of the Calculus\label{sec:sound}}
Two aspects determine the soundness of the calculus: a closed branch must be unsatisfiable and all expansion rules must maintain satisfiability.
We can express the former aspect in Isabelle as follows.
\begin{lstlisting}[belowskip=0pt,label={lst:bclosed_sound}]
lemma bclosed_sound:
  assumes bclosed b shows $\exists$$\phi$ $\in$ set b. M $\not\models$ $\phi$
\end{lstlisting}
\begin{proof}
  Assume for contradiction that \lstinline!b! is closed but we have a model, i.e.\ it holds that \lstinline!$\forall$$\phi$ $\in$ set b. M $\models$ $\phi$!.
  We perform a case analysis on \lstinline!bclosed b!.
  It is clear that, for any \lstinline!s!, neither does \lstinline!M! model \lstinline!s $\in$ $\emptyset$! nor \lstinline!s $\neqS$ s!.
  Furthermore, no model can satisfy both \lstinline!$\phi$! and \lstinline!&\fmNeg{$\phi$}&! at the same time.
  This leaves us with the case where \lstinline!b! contains a membership cycle
\begin{lstlisting}
  t$_\text{0}$ $\inS$ t$_\text{1}$, t$_\text{1}$ $\inS$ t$_\text{2}$, $\ldots$, t$_\text{k}$ $\inS$ t$_\text{0}$.
\end{lstlisting}
  By assumption, each atom of the membership cycle holds under the model \lstinline!M!; in other words, we have
\begin{lstlisting}
  &\Ist& M t$_\text{i}$ $\in$ elts (&\Ist& M t$_\text{i + 1 mod k + 1}$)
\end{lstlisting}
  for each \lstinline!i $\in$ {0,$\ldots$,k}!.
  Therefore, the transitive closure of the membership relation \lstinline!{(x, y). x $\in$ elts y}! contains a cycle. 
  This, however, is a contradiction since the membership relation of \lstinline!V! is well-founded (see \autoref{sec:semantics}) and well-foundedness is closed under transitive closure.
\end{proof}

\noindent We are left with showing that both linear and branching expansion rules preserve satisfiability.
As for the linear rules, a straightforward proof by case analysis on \lstinline!&\lexpands{b'}{b}&! suffices to obtain the lemma below.
\begin{lstlisting}[label={lst:lexpands_sound}]
lemma lexpands_sound:
  assumes &\lexpands{b'}{b}& and $\phi$ $\in$ set b'
  assumes $\bigwedge$$\psi$. $\psi$ $\in$ set b $\Longrightarrow$ M $\models$ $\psi$
  shows M $\models$ $\phi$
\end{lstlisting}
A similar argument would work for the branching rules if it were not for the last rule adding new variables.
Those variables need to be assigned specific values; hence, we modify the model as we will see in the proof below.
\begin{lstlisting}[belowskip=0pt, label={lst:bexpands_sound}]
lemma bexpands_sound:
  assumes &\fexpands{bs'}{b}&
  assumes $\bigwedge$$\psi$. $\psi$ $\in$ set b $\Longrightarrow$ M $\models$ $\psi$
  shows $\exists$M'. $\exists$b' $\in$ bs'.
          $\forall$$\psi$ $\in$ set (b' @ b). M' $\models$ $\psi$
\end{lstlisting}
\begin{proof}
  We perform a case analysis on the derivation \lstinline!&\fexpands{bs'}{b}&!.
  As all cases but that of the last rule are straightforward, we only consider this last case where we have 
\begin{lstlisting}
  bs' = {[Var x $\inS$ s, Var x $\notinS$ t],
         [Var x $\inS$ t, Var x $\notinS$ s]}
\end{lstlisting}
  for some fresh variable \lstinline!x!, i.e.\ \lstinline!x $\notin$ vars b! and in particular \lstinline!x $\notin$ vars s $\cup$ vars t!.
  Another consequence of the last rule being applied is that \lstinline!s $\neqS$ t! is in \lstinline!b!, which, because \lstinline!M! is a model, implies that \lstinline!&\Ist& M s $\neq$ &\Ist& M t!.
  This inequality manifests itself through some \lstinline!y! with
\begin{lstlisting}
  &\phantom{$\lor$}& y $\in$ elts (&\Ist& M s) $\land$ y $\notin$ elts (&\Ist& M t)
  $\lor$ y $\in$ elts (&\Ist& M t) $\land$ y $\notin$ elts (&\Ist& M s).
\end{lstlisting}
  We now update the assignment \lstinline!M! such that it maps \lstinline!x! to \lstinline!y! to obtain the assignment \lstinline!M'!.
  Note that values of both \lstinline!s! and \lstinline!t! are unchanged under this update as \lstinline!x $\notin$ vars s! and \lstinline!x $\notin$ vars t!.
  Considering our assumption about \lstinline!y! and the fact that \lstinline!&\Ist& M' (Var x) = y!, we infer that
\begin{lstlisting}
  $\exists$b' $\in$ bs'. $\forall$$\psi$ $\in$ set b'. M' $\models$ $\psi$.
\end{lstlisting}
  In other words, due to the disjunction, it holds that \lstinline!M'! is a model for one \lstinline!b' $\in$ bs!.
  Thus, we are left with proving that \lstinline!M'! is a model for \lstinline!b!.
  This is indeed the case since the assumption \lstinline!M $\models$ $\psi$! is equivalent to \lstinline!M' $\models$ $\psi$! for all \lstinline!$\psi$ $\in$ set b! given that \lstinline!x! is fresh with respect to b.
\end{proof}
